Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-12-11 text/ms.tex
Minor edits to main text

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, estimate the degree of
heavy-tailedness (presence of black swans) in ecological process noise by
applying a probability model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

@@

Across populations, the probability of observing black-swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Figs~\ref{fig:correlates}, \ref{fig:posteriors}b). Longer time-series length
was the strongest covariate of observing black-swan dynamics. For instance, we
were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a
population with 60 time steps compared to one with 30 time steps. However, the
absolute change in probability with increased time series length was small
($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example,
Fig.~\ref{fig:correlates}).

@@

We might expect to observe black-swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions \citep[with the same mean and inverse-gamma-distributed
variances,][]{gelman2014}, we could observe heavy tails if we miss some
underlying mixture of intrinsic processes \citep{allen2001}. That process
might be an aggregation of populations across space, or population diversity,
or an intrinsic change in population variability through time. Extrinsic
forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These
forces could be extreme themselves. For example, extreme climate, predation
from (or competition with) other species experiencing black swans, or sharp
changes in human pressure such hunting, fishing, or habitat destruction might
cause black swans. Alternatively, the synchrony of multiple ``normal''
extrinsic forces could give rise to black-swan ecological dynamics. This could
occur if with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
non-synergistic forces experience a rare alignment \citep{denny2009}.

@@

Most importantly, what is the impact of allowing for black swans in forecasts
of ecological risk? Even extremely rare catastrophes can have a profound
influence on population persistence \citep{mangel1994}. In recent decades,
ecology has moved toward focussing on aspects of variance in addition to mean
responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that
an added focus on ecological extremes represents the next frontier,
particularly in the face of increased climate extremes \citep{meehl2004,
  thompson2013, ipcc2012}. Financial analysts are concerned with the shape of
the downward tails of financial returns because these directly impact
estimates of risk---the probability of a specific magnitude of undesired event
occurring \citep{rachev2008}. A comparable focus in ecology would increase our
estimates of extinction risk, since these would be disproportionately impacted
by downward black-swan events.

@@

We thank J.D. Yeakel and other members of the Earth to Ocean research
group for helpful discussions and comments. We are grateful to the
contributors and maintainers of the Global Population Dynamics Database and to
Compute Canada's WestGrid high-performance computing resources. Silhouette
images were obtained from \texttt{phylopic.org} under Creative Commons
licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
Research Chairs Program (NKD).

>>>>

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, we estimate the degree of
heavy-tailedness (presence of black swans) in ecological process noise by
applying a probability model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%) and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

@@

Across populations, the probability of observing black-swan dynamics was
positively related to time-series length and negatively related to magnitude
of process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Figs~\ref{fig:posteriors}b, \ref{fig:correlates}). Longer time-series length
was the strongest covariate of observing black-swan dynamics. For instance,
the expected probability density below $\nu = 10$ was about \pIncHeavyNThirtyNSixty~times
greater for a population with 60 time steps compared
to one with 30 time steps. However, the absolute change in probability with
increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$
in the previous example, Fig.~\ref{fig:correlates}).

@@

We might expect to observe black-swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions \citep[with the same mean and inverse-gamma-distributed
variances,][]{gelman2014}, we could observe heavy tails if we miss some
underlying mixture of intrinsic processes \citep{allen2001}. That process
might be an aggregation of populations across space, or population diversity,
or an intrinsic change in population variability through time. Extrinsic
forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These
forces could be extreme themselves. For example, extreme climate, predation
from (or competition with) other species experiencing black swans, or sharp
changes in human pressure such hunting, fishing, or habitat destruction might
cause black swans. Alternatively, the synchrony of multiple ``normal''
extrinsic forces could give rise to black-swan ecological dynamics. This could
occur with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
non-synergistic forces experience a rare alignment \citep{denny2009}.

@@

Most importantly, what is the impact of allowing for black swans in forecasts
of ecological risk? Even extremely rare catastrophes can have a profound
influence on population persistence \citep{mangel1994}. In recent decades,
ecology has moved toward focussing on aspects of variance in addition to mean
responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that
an added focus on ecological extremes represents the next frontier,
particularly in the face of increased climate extremes \citep{meehl2004,
  ipcc2012, thompson2013}. Financial analysts are concerned with the shape of
the downward tails of financial returns because these directly impact
estimates of risk---the probability of a specific magnitude of undesired event
occurring \citep{rachev2008}. A comparable focus in ecology would increase our
estimates of extinction risk, since these would be disproportionately impacted
by downward black-swan events.

@@

We thank J.W. Moore, J.D. Yeakel, and other members of the Earth to Ocean
research group for helpful discussions and comments. We are grateful to the
contributors and maintainers of the Global Population Dynamics Database and to
Compute Canada's WestGrid high-performance computing resources. Silhouette
images were obtained from \texttt{phylopic.org} under Creative Commons
licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
Research Chairs Program (NKD).
